Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-03-24 article.tex
Progress on results

While assuming that an earthquake's magnitude is independent of the seismic history is popular due to its simplicity and its ease of use in earthquake modeling, it is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field. These in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability, i.e. its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools. Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding. 
\newline
% perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's
In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history. This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

@@

Quantification of a model's preformance in our scheme will be done by averging on all log-likelihood scores of the test set, as defined in \ref{eq:loss_function}. \ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from \ref{eq:loss_function}, for MAGNNET and chosen common benchmarks, for the three regions tested. It be clearly seen that MAGNNET shows a clear an significant average information gain over every benchmark tested. \neri{Should we mention other benchmarks and refer to the SM?}.

@@

As the out model's output is a probability density function (PDF), we may easily convert it to a binary classifer. By doing this we are posing a modified question: will there be a large earthquake in the query space-time coordiantes or not? This is easily done by setting a threshold which defines a "large" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative. The probability of a large event occuring in the query coordinates is given by integrating the output PDF above the threshild, $p=\int_{m_c}^{\infty}p(m')dm'$. Given such binary classifier, it is possible now to examine the common binary classifier metrics of area under reciever operator (ROC) curve [cite] and the prescision at recall curve [cite]. \ref{fig:metrics}d-f show the ROC curves for the binary modification for all three regions in question. The area under the curve (AUC) are written to the bottom right of each panel color coded by the same legend as the bar plots above. It can be seen that throughout all regions MAGNNET preforms significantly better than all other benchmarks, indicated by the largest AUC value. The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNNET's curve displays the largest AUC among all benchmarks tested throughout all regions.

@@

The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
\begin{equation}
    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(model)}(m_i)} - \log{p^{(GR)}(m_i)}
    \label{eq:information_gain}
\end{equation}
This equation measures the number of bits of information gained by using our model compared to the GR benchmark. Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage. \ref{fig:info_gain_over_time}a shows the test set for Southern California event magnitudes by their running index, colored according to the information gain of our model compared to the GR benchmark. This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as a black dashed line in the figure. The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods. Notably, for the Southern California data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes. Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4$ bits, respectively, though these events are typically considered completely unpredictable. This observation leads to a significant corollary: at least some portion of the information gain likely originates from a physical precursory signal. This point will be further elaborated upon in the discussion section.
\newline
The observed increase in information gain persists throughout the entire test set for all regions examined in this study. This includes regions previously analyzed by (Ogata et al. SRL 2018), where no information gain was detected using the method employed at the time. Further analysis of \ref{fig:likelihood_over_time} applied to other regions is presented in the supplementary material.
\newline
While the information gain observed in pre- and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)? To address this, we factor out this potential bias by recalculating the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:

@@

For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method (Weimer and Weiss, 2000) within a window of 150 past and 150 future events. The resulting temporal incompleteness over time is displayed in \ref{fig:likelihood_over_time}b.
Crucially, \ref{fig:likelihood_over_time}b also presents the magnitudes of the Ridgecrest sequence at their origin times, colored by the information gain of our model over the GR benchmark, both conditioned on values above $m_c(t)$ as in \ref{eq:conditioned_likelihood}. This figure clearly demonstrates that even after accounting for temporal incompleteness, the information gain achieved by our model remains an increasing function throughout the test set. The mean score of the test set for the conditioned results are shown in [[]] which displays an outpreformance for our model with this metric as well. This finding underscores the ability of our model to capture true physical features beyond mere measurement artifacts.

[TODO: should I keep of remove the metrics for the causality defying metrics (present events) that beat the model in the ROC? this affects how the rest of this chapter will be written]

>>>>

While assuming that an earthquake's magnitude is independent of the seismic history is popular due to its simplicity and its ease of use in earthquake modeling, it is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field. These in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability, i.e. its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools. Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding. 
\newline
\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's}
In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history. This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

@@

Quantification of a model's preformance in our scheme will be done by averging on all log-likelihood scores of the test set, as defined in Eq. \ref{eq:loss_function}. Fig. \ref{fig:metrics}a-c shows the minus mean log likelihood score, $\mathcal{L}$ from Eq. \ref{eq:loss_function}, for MAGNNET and chosen common benchmarks, for the three regions tested. It can be clearly seen that MAGNNET shows a clear an significant average information gain over every benchmark tested. \neri{Should we mention other benchmarks and refer to the SM?}.

@@

As the out model's output is a probability density function (PDF), we may easily convert it to a binary classifer. By doing this we are posing a modified question: will there be a \textit{large} earthquake in the query space-time coordiantes or not? This is done by setting a threshold which defines a "\textit{large}" earthquake, $m_c$ ($=4$ in the examples presented here), then setting examples above that threshold as positive examples and below as negative. The probability of a large event occuring in the query coordinates is given by integrating the output PDF above the threshild, $p=\int_{m_c}^{\infty}p(m')dm'$. Given such binary classifier, it is possible now to examine the common binary classifier metrics of area under reciever operator (ROC) curve [cite] and the prescision at recall curve [cite]. \ref{fig:metrics}d-f show the ROC curves for the binary modification for all three regions in question. The area under the curve (AUC) are written to the bottom right of each panel color coded by the same legend as the bar plots above. It can be seen that throughout all regions MAGNNET preforms significantly better than all other benchmarks, indicated by the largest AUC value. The precision at recall curve displayed in \ref{fig:metrics}g-i show a similar result as MAGNNET's curve displays the largest AUC among all benchmarks tested throughout all regions.
\neri{TODO: calculate and present AUC P@R}

@@

The information gain achieved by our model for each event can be quantified by the difference in log-likelihood (LL) between our model's prediction and the baseline GR predictor:
\begin{equation}
    \Delta LL = \log{p_{\pmb{x_i}, t_i}^{(MAGNNET)}(m_i)} - \log{p^{(GR)}(m_i)}
    \label{eq:information_gain}
\end{equation}
This equation measures the amount of information gained by using our MAGNNET model compared to the GR benchmark. Analyzing this quantity per event across the entire test set reveals patterns in where our model achieves its advantage. Fig. \ref{fig:info_gain_over_time}a shows the test set for Southern California event magnitudes by their running index, colored according to the information gain of our model over the GR benchmark. This time-domain representation allows us to track the cumulative information gain, calculated over the entire test set and displayed as the black dashed line in the figure. The predominantly increasing trend in cumulative information gain suggests that our model is consistently advantageous as opposed to a concentrated advantage in specific periods. Notably, for the Southern California data set, the most rapid information gain occurs during the aftershocks of the two Ridgecrest earthquakes. Surprisingly, even the pre-shock and main shock events demonstrate positive information gains over the GR benchmark, with values of $0.5$ and $1.4, \ \Delta LL$, respectively, albeit these events are typically considered completely unpredictable. This observation leads to a significant corollary: at least some portion of the information gain likely originates from a statistical signal embedded in the event sequence prior to event's occurence. This point will be further elaborated upon in the discussion section.
\newline
The observed increase in information gain persists throughout the entire test set for all regions examined in this study as the black dashed curves in Fig. \ref{fig:info_gain_over_time}b-d present. This includes regions previously analyzed by (Ogata et al. SRL 2018), where no information gain was detected using the method employed at the time. Further analysis of \ref{fig:info_gain_over_time} applied to other regions is presented in the supplementary material. \neri{TODO: ad these colorful plots for NZ and JAPAN to the SM}
The smoothed instantaneous information gain per event is presented in Fig. \ref{fig:info_gain_over_time}e-g, and can be seen as commonly well above the threshold of zero information gain, marked by the red dashed line.
\newline
While the information gain observed in preshock and mainshocks presented above is encouraging, a potential concern might arise regarding the aftershocks: could their information gain be an artifact of fitting to the temporal incompleteness of the catalog (cite somethong?)? To address this, we factor out this potential bias by recalculating the likelihood score of each event, $p(m_i)$ , conditioned on the magnitude exceeding a threshold, $\Tilde{m}$:

@@

For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method (Weimer and Weiss, 2000) within a window of 150 past and 150 future events. The resulting temporal incompleteness curves are presented in the SM \neri{TODO: add the $m_c(t)$ figure to SM}. After applying the renormalization of Eq. \ref{eq:conditioned_likelihood} the cumulative information gain is recalculated. The result is presented as the grey dashed curved in Fig.  \ref{fig:info_gain_over_time}a, as well as in Fig. \ref{fig:info_gain_over_time}b-d for all three tested regions.
\newline
A second concern regarding temporal variations of the completeness magnitude may be raised to oppose to the analysis and results presented in this work. The following paragraph will provide an explanation of our method for varifying that we are indeed measuring a true statistical trend that is not created by the local completeness magnitude artifact. We start by deviding each region into bins of size [dx, dy] \neri{TODO fill in the details}. The average log likelihood score $\mathcal{L}$ is calculated per bin for both MAGNNET and GR, resulting in a map of spatial average log likelihood, per model. The subtraction of the two maps is presented in Fig. \ref{fig:info_gain_over_time}h-j, showing the spatial distribution of information gain for each region. 
\neri{TODO complete paragraph about how I select only bins of mc(x,y)<=mc to recalculate the average score and show that I still gain information.}
for which the local completness magnitude $m_c(x, y)$ is calculated.
